\chapter{State of the Art}
\label{chapter:sota} 

The SDN Product Directory\cite{onf:products} page from Open Networking Foundation\cite{onf} lists
several products for hardware switching and virtual switching. Of all the virtual switching solutions,
only two of them mention supporting OpenFlow version 1.3. However, these are proprietary products that
we can't contribute to.

Another solution for virtual switching that this time is open source is Open vSwitch. It is able to connect to OpenFlow controllers,
but it does not support the newest additions to the protocol. The documentation mentions that 
``support for OpenFlow 1.1 and beyond is a work in progress''\footnote{\url{https://github.com/openvswitch/ovs/blob/master/OPENFLOW-1.1\%2B}}.
We can use Open vSwitch as a base of development for implementing our project.
 
In order to work on new features for Open vSwitch, we must also analyze existing solutions for implementing controllers
that use OpenFlow 1.4. We want to focus on the switch implementation so we don't want to write a controller from
scratch. We present in the next section the options that we have.
 
\section{Related Work}

% TODO: Floodlight

On the controller side, there are also several implementations of frameworks used for developing
various network management and control applications. POX\cite{pox} can be used to write various networking software,
including OpenFlow switches and controllers. However, it can only communicate with OpenFlow 1.0 switches.
Adding support for newer versions is out of the scope of this project.

One of the relatively new SDN projects is OpenDaylight\footnote{\url{http://www.opendaylight.org/}}. It is lead by
the Linux Foundation and aims to improve the development and adoption of SDN. Currently, they offer a modular controller,
written in Java. Modules are provided both for northbound and southbound APIs. Northbound APIs are used by high level
network application to implement business logic. The southbound APIs allow the controller to act on the physical or
virtual network devices, like switches. The support for OpenFlow is, however, limited to versions 1.0 and 1.3, so
we had to find other options.

Another SDN framework is Ryu\cite{ryu}. It already supports various management protocols, including OpenFlow up to 1.4.
This can be a useful tool to validate our work on Open vSwitch. We will use it to create a simple learning switch
controller that can also send OpenFlow 1.4 bundle messages.

\section{Project Motivation}

Keeping a consistent state between multiple switches and controllers is challenging. Configuration
changes can happen at either of them and the separation paradigm introduced by SDN makes it even
more difficult. For example, we might need to propagate changes to several devices. These changes form
a logical group and are not useful if not applied together. So one failure at this stage is already too
late. That is why we need to keep the controller informed of the status of the switch (by notifying it of
permissions, available resources, various changes) and allow it to form logical groups of messages that
must be executed together.

One use case of batching OpenFlow messages using bundles is to have synchronized changes across multiple switches,
for example, during an initialization phase. The pre-validation of messages and atomic commit ensure
that the state remains consistent between the controller and several switches.

Another use case for batching messages is that of network security application controllers. These can range
from simple stateless firewalls to intrusion detection systems\cite{sdn}. They work by installing flows to
allow or drop future packets. The decision to do that can be static, based on access control list, or dynamic,
by inspecting the traffic patterns to determine the application types and permissions.
These types of security solutions can be built on top of OpenFlow controllers. By batching OpenFlow
messages we can ensure that security rules are enforced correctly and completely. In addition, we can
deploy a set of rules to multiple devices at the same time.

